---
title: "Elements of Data Science <br> SDS 322E"
author: "H. Sherry Zhang <br> Department of Statistics and Data Sciences <br> The University of Texas at Austin <br>  <br> week 01 lecture 01 Fall 2025"
format: 
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    aspectratio: 169
    theme: serif
    preview-links: auto
    pdf-separate-fragments: true
---

planned:

- week 1: intro
- week 2: some buffer: some base R, download data, read in data: `readr`, `janitor`, summary
- week 3 & 4: data wrangling: `tidyr`, `dplyr`, `stringr`, `lubridate`
- week 5 & 6: visualization: `ggplot2`: more nitty gritty
- week 7: some buffer? 
- week 8: text data



- week 14: spatial data

buffer topics:

- some base R doesn't hurt
- web scraping
- writing functions + functional programming

# [Example: search and install from GitHub]{.r-fit}

I want to add emoji in the lecture slides and I know there is the `emo` package. The syntax is `emo::ji("emoji_name")`. 

```{r eval = FALSE}
emo::ji("evil")
```

> Error in loadNamespace(x) : there is no package called ‘emo’

Okay... I don't have it in my lastest R, let's install it...

```{r eval = FALSE}
install.packages("emo")
```

> Warning in install.packages :
  package ‘emo’ is not available for this version of R
  A version of this package for your version of R might be available elsewhere, see the ideas at https://cran.r-project.org/doc/manuals/r-patched/R-admin.html#Installing-packages

Aha - it's Hadley's GitHub package, let me find it: 

Google: emo ji Hadley github

and find this: https://github.com/hadley/emo

Here I find the code and Let's run this: 

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("hadley/emo")
```


now 

```{r}
emo::ji("devil")
```

